---
title: "TITLE OF YOUR PROJECT"
subtitle: "Exploring Disparate Data: Part 3 - Final Report"
author: "Put your group name here: Liza&Maya"
date: "Due November 26th, 2025"
output: pdf_document
---

List your group members, including their student numbers, here:

-   Maya McNeill (\)
-   Liza Luizova (169089763)

### How do Arctic environmental changes relate to global patterns of well being and climate awareness? 

# Data Description

## \<\Sea Ice extent data\>\>

```{r load_data1}
library(tidyverse)
library(arrow)
library(lubridate)
# Put in your code to load in the data set, along with any
# necessary cleaning beyond what was done in Part 1

# Note that the code in this document will not be shown
# when you click "knit", so the placement of this code
# chunk is purely for your benefit: You can see what happened
# with your data, which makes it easier to describe below!
ice_extent <- read_parquet("ice_extent_yeary.parquet") |>
  filter(region == "Arctic") |>
  pivot_wider(
    names_from = name,
    values_from = value
  ) |>
  mutate(ice_loss = max - min) |>
  select(year, arctic_max_ice = max, arctic_min_ice = min, ice_loss)
ice_extent

```

The data is obtained from the National Oceanic and Atmospheric Administration (NOAA), the published Arctic report card for sea ice extent from 1978 to 2023. 

In order to clean the data, we filtered the Arctic region measurements, pivoted to wide format to create separate columns for maximum and minimum ice extent, calculated annual ice loss as the difference between maximum and minimum, and selected relevant columns for analysis, such as year, the change in ice(ice_loss), and maximum/minimum amount of ice across years(arctic_max_ice, arctic_min_ice).  

## \<\<Climate awareness data set\>\>

```{r load_data2}
# Put in your code to load in the data set, along with any
# necessary cleaning beyond what was done in Part 1

# Reminder: do NOT print your data to the screen unless it's
# completely necessary
climate_awareness <- read_parquet("climate_awareness.parquet") |>
  mutate(
    high_awareness = aware_alot + aware_moderate,
    climate_awareness_index = (aware_alot * 1 + aware_moderate * 0.67 + 
                               aware_alittle * 0.33) / 100
  ) |>
  select(country, high_awareness, climate_awareness_index, 
         aware_no, aware_alittle, aware_moderate, aware_alot)
climate_awareness
```

The data is obtained from Meta's Climate Change Opinion Survey 2022 and detail public awareness levels about climate change definitions across countries.

In order to clean the data, we created composite measures including a high awareness percentage (combining "a lot", "moderate" and "a little" awareness) and a weighted climate awareness index to capture the spectrum of awareness levels.Then we selected the relevant columns, such as levels of awareness across countries, that are going to be used for our further exploration.  

## \<\<Happiness data set\>\>

```{r load_data3}
# Put in your code to load in the data set, along with any
# necessary cleaning beyond what was done in Part 1
happiness <- read_parquet("happiness.parquet") |>
  select(country, year, happiness_score = ladder_score, 
        gdp_contribution = explained_by_log_gdp_per_capita, 
         social_support = explained_by_social_support, 
         life_expectancy = explained_by_healthy_life_expectancy,
         freedom = explained_by_freedom_to_make_life_choices, 
         generosity = explained_by_generosity, 
         corruption = explained_by_perceptions_of_corruption) |>
  filter(!is.na(happiness_score))
happiness
```

The data come from \<<place>\> and detail \<<more specific description of the data>\>.

In order to clean the data, we \<\<steps to clean the data, concise but precise enough that a reader could follow your steps without seeing your code\>\>

## Combining the Data

We combined the datasets by joining on country and year where applicable. For the sea ice data, which represents global environmental conditions, we joined with country-level data by year to examine how Arctic changes correlate with national outcomes. Country name standardization was required, particularly for the climate awareness data where we matched country names to the happiness report naming conventions using case_when statements for manual recoding of discrepancies.

# Exploratory Data Analysis

To achieve our goals, we explored the data by examining temporal trends in Arctic ice loss, correlating these changes with national happiness metrics, and investigating whether climate awareness moderates these relationships across different geographic regions.

We explored many aspects of the data, but will demonstrate three. These are the relationship between recent Arctic ice loss and national happiness, how climate awareness varies with proximity to the Arctic, and whether countries with higher climate awareness show different patterns in the ice loss-happiness relationship.

The first aspect that we found interesting is shown in @ref(fig:trend_analysis). Countries experiencing the most dramatic Arctic ice loss years show varied happiness responses, with northern countries displaying different patterns than tropical nations.

```{r figure1, fig.cap="Parallel trends in the arctic loss and the average happiness score from 2011-2021, showing potential environmental impacts on global wellbeing."}
# This is an example of how you can control figures and captions in
# an R chunk. Note that you can reference figures using:
# \@ref(fig:insight1), where "insight1" is the label of this code
# chunk (the first bit of text after the "r" in "```{r label, options...}")
library(patchwork)
cat("Happiness data:\n")
print(head(happiness))
cat("Years in happiness:", range(happiness$year, na.rm = TRUE), "\n")

cat("\nClimate awareness data:\n") 
print(head(climate_awareness))

cat("\nIce extent data:\n")
print(head(ice_extent))
cat("Years in ice extent:", range(ice_extent$year, na.rm = TRUE), "\n")

happiness_years <- unique(happiness$year)
ice_years <- unique(ice_extent$year)
common_years <- intersect(happiness_years, ice_years)
cat("\nCommon years:", common_years, "\n")

combined_simple <- happiness |>
    filter(year %in% common_years) |>
    group_by(year) |>
    summarise(avg_happiness = mean(happiness_score, na.rm = TRUE)) |>
    left_join(ice_extent, by = "year")

p1 <- ggplot(combined_simple) +
    aes(x = year, y = avg_happiness) +
    geom_line(color = "blue", size = 1) +
    geom_point(color = "red", size = 2) +
    labs(x = "Year", y = "Average Happiness Score",
         title = "Average Happiness Over Time") +
    theme_minimal()
  
  p2 <- ggplot(combined_simple) +
    aes(x = year, y = ice_loss) +
    geom_line(color = "red", size = 1) +
    geom_point(color = "blue", size = 2) +
    labs(x = "Year", y = "Ice Loss (million km²)",
         title = "Arctic Ice Loss Over Time") +
    theme_minimal()
  p1 / p2
```

This insight is supported by the summary statistics in table @ref(tab:summary_stats)

```{r summary_stats}
# Calculate the relevant summary statistics here.
# Note that the "kable" function in the "knitr" package
# is convenient for making nice tables. Other packages can
# do much fancier things with tables, but keep in mind that
# the insights should be the star, not the formatting.
library(knitr)

summary_table <- happiness |>
  inner_join(ice_extent, by = "year") |>
  filter(year >= 2011 & year <= 2021) |>  # Match your plot timeframe
  group_by(year) |>
  summarise(
    avg_happiness = mean(happiness_score, na.rm = TRUE),
    avg_ice_loss = mean(ice_loss, na.rm = TRUE),
    n_countries = n_distinct(country),
    happiness_sd = sd(happiness_score, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(across(where(is.numeric), round, 3))

kable(summary_table, caption = "Yearly summary statistics for happiness and ice loss (2011-2021)")
```
The next insight that we found is shown in @ref(fig:awareness_geography). Northern countries closest to the Arctic show both higher climate awareness and more variable responses to environmental changes.


```{r insight2, fig.height=4, fig.width=6, fig.cap="This is a figure caption that you will need to change in order to get good marks in the visualization rubric items."}
# This figure will have a height of 4 and a width of 6.
# Feel free to change this, and to apply different sizes
# to the other figures you create.
awareness_summary <- combined_data |>
  group_by(region, country) |>
  summarise(avg_awareness = mean(climate_awareness_index, na.rm = TRUE),
            .groups = "drop")

ggplot(awareness_summary) +
  aes(x = region, y = avg_awareness, fill = region) +
  geom_violin(alpha = 0.6, show.legend = FALSE) +
  geom_jitter(width = 0.2, alpha = 0.7, size = 1.5) +
  labs(x = "Geographic Region",
       y = "Climate Awareness Index",
       title = "Climate Awareness Distribution by Geographic Region") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Finally, @ref(fig:time_evolution) shows how the relationship between Arctic changes and global wellbeing has evolved over time, with climate awareness potentially playing a protective role in recent years.


```{r insight3, fig.height=4, fig.width=6, fig.cap="This is a figure caption that you will need to change in order to get good marks in the visualization rubric items."}
# Create correlation over time analysis
# Create a high happiness indicator first
# Simple and robust approach
# Safe correlation calculation with data sufficiency check
yearly_cor <- combined_data |>
  group_by(year) |>
  filter(sum(complete.cases(ice_loss, happiness_score)) >= 5) |>  # Require sufficient data
  summarise(
    correlation = cor(ice_loss, happiness_score),
    n_countries = sum(complete.cases(ice_loss, happiness_score)),
    .groups = "drop"
  ) |>
  filter(year >= 2000, !is.na(correlation))

ggplot(yearly_cor) +
  aes(x = year, y = correlation) +
  geom_col(aes(fill = correlation > 0), alpha = 0.7, show.legend = FALSE) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(x = "Year", 
       y = "Correlation Coefficient",
       title = "Yearly Correlation: Arctic Ice Loss vs National Happiness",
       subtitle = "Positive values indicate happier countries during years of greater ice loss") +
  scale_fill_manual(values = c("FALSE" = "#E74C3C", "TRUE" = "#3498DB")) +
  theme_minimal() +
  scale_x_continuous(breaks = seq(2000, 2023, by = 3))
```

# Conclusion and Future Work

Overall, we found \<<general ideas>\>.

A second paragraph about our findings.

The next steps in this analysis are...

The limitations of this analysis are as follows. (Do not simply list potential issues with sampling, but relate them to your analysis and how they affect your conclusions. An honest and complete acknowledgement of the limitations makes the analysis more trustworthy.)

# References

I am not strict about MLA or APA style or anything like that. For this report, I would much rather have your citations be easy to match to your insights.

The easiest way is to use Rmd's [footnote](https://bookdown.org/yihui/rmarkdown/markdown-syntax.html#inline-formatting) syntax. This will put a number beside the word where the footnote appears, and the full text of the footnote at the bottom of the page (pdf) or end of the document (html). The syntax is:[^1], where I suggest that you put in something like this[^2] to make references for this assignment.

[^1]: See the source view to see this footnote

[^2]: The relevance to the insight is ... . From \<<name of source and name of article>\>, published on \<<date>\>, url: \<<link to page>\>

Alternatively, you could make a list of citations with their main arguments and why they're relevent to your insights, methods, etc.

The link above also references "bibtex" files. These are also extremely convenient, but have a steep learning curve and they make it difficult to tie them to an insight. If you use bibtext, then make sure that you provide a sentence to describe the source and it's relevance when you cite it - don't just add citations to the end of a sentence (this is common practice in academia, but I want to know that your citations are directly relevant for this assignmnet).
